### 2014-05-12 writeups/context-dependence.tex
Confusion regarding the starting state.
@petrelharp-- could you please let me know if I'm off the mark with this
commit? Is X(0) drawn from a distribution or can we start it with an
arbitrary series of states?
A little clarification on my confusion.

The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
The sequence-level transition process is the independent superposition of all of the individual pattern mutation processes.
Thus given two patterns $u$ and $v$ of common length $h$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+h}(t)) = (u_1, \ldots, u_h)$ then
\[
\P\{(X_i(t+dt), \ldots, X_{i+h}(t+dt)) = (v_1, \ldots, v_h) = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
We call such a $(\mu,u,v)$ a ``transition triple''.
The set of all transition triples $\calT = \{ (\mu^j,u^j,v^j) \}_{j=1}^{n_T}$
determines the time evolution of the process.

@@

Now we will now calculate the expectation and covariances of $N_\ell(x,y)$.
From the definitions above,
\[
\E [N_\ell(x,y)] \approx N(x) \, p_{\lyl}(x,y).
\]
To compute covariances, define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$.
\begin{align*}
  \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^{n-1} \cov[M_1(x,y),M_{k+1}(x,y)] \\
  &\approx n p_{\lyl}(x,y)(1-p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2 \right) \\
  &= n p_{\lyl}(x,y)(1-(1+2m+4\ell)p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
\end{align}
The term $\E[M_1(x,y) M_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
\begin{align}
  \begin{split}
    \cov[N(x_1,y_1),N(x_2,y_2)]
    & \approx - n (2m+4\ell) p_{\lyl}(x_1,y_1)p_{\lyl}(x_2,y_2) \\
    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[M_1(x_1,y_1) M_{k+1}(x_2,y_2) + M_1(x_2,y_2) M_{k+1}(x_1,y_1)] .
  \end{split}
\end{align*}

>>>>

The process starts with an arbitrary specified sequence of states $X(0)$.
The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
The sequence-level transition process is the independent superposition of all of the individual pattern mutation processes.
Thus given two patterns $u$ and $v$ of common length $h$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+h}(t)) = (u_1, \ldots, u_h)$ then
\[
\P\{(X_i(t+dt), \ldots, X_{i+h}(t+dt)) = (v_1, \ldots, v_h) = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
We call such a $(\mu,u,v)$ a ``transition triple''.
The set of all transition triples $\calT = \{ (\mu^j,u^j,v^j) \}_{j=1}^{n_T}$
determines the time evolution of the process.

@@

Now we will now calculate the expectation and covariances of $N_\ell(x,y)$.
From the definitions above,
\[
\E [N_\ell(x,y)] \approx N(x) \, p_{\lyl}(x,y).
\]
To compute covariances, define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$.
\todo[inline]{
In the below, are you are using an equivalence
\[
\cov[M_1(x,y),M_{k+1}(x,y)] = \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2
\]
How do you justify this? I don't think that $M_1(x,y)$ is equal to $p_{\lyl}(x,y)$: the former is defined in terms of observed events, whereas the latter is defined starting with "if we have a sequence $x$" without any reference to the frequency of $x$.
For example, what if $x$ doesn't appear in $X(0)$?
I was with you on $\E [N_\ell(x,y)] \approx N(x) \, p_{\lyl}(x,y)$.
}

\begin{align*}
  \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^{n-1} \cov[M_1(x,y),M_{k+1}(x,y)] \\
  &\approx n p_{\lyl}(x,y)(1-p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2 \right) \\
  &= n p_{\lyl}(x,y)(1-(1+2m+4\ell)p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
\end{align}
The term $\E[M_1(x,y) M_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
\begin{align}
  \begin{split}
    \cov[N(x_1,y_1),N(x_2,y_2)]
    & \approx - n (2m+4\ell) p_{\lyl}(x_1,y_1)p_{\lyl}(x_2,y_2) \\
    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[M_1(x_1,y_1) M_{k+1}(x_2,y_2) + M_1(x_2,y_2) M_{k+1}(x_1,y_1)] .
  \end{split}
\end{align*}
